Visual narratives of the Covid-19 pandemic

### 2022-01-15 JDSSV_manuscript_knitr.Rnw
Some more log stuff

@@

\begin{document}
% \SweaveOpts{concordance=TRUE}
<<include=FALSE, cache=FALSE, echo=FALSE, purl=FALSE>>=
library(knitr)
opts_chunk$set(comment = NA, size = 'footnotesize', prompt = TRUE, highlight = FALSE, 
               cache = TRUE, crop = TRUE, concordance = FALSE, fig.align='center',
               fig.path='Figures_R/fig-',out.width="0.4\\textwidth")
opts_knit$set(out.format = "latex")
knit_hooks$set(crop = hook_pdfcrop)
options(width = 80, prompt = "R> ", continue = "    ", digits = 3)
thm <- knit_theme$get('default')
Sys.setenv(TEXINPUTS=getwd(),
           BIBINPUTS=getwd(),
           BSTINPUTS=getwd())
knit_theme$set(thm)
@

@@

<<freq-domain-plot, include = F, eval = F>>
show_states <- c("California", "Texas", "New York", "Arizona", "Iowa", "Washington")

@@

\section{To log or not to log}
As COVID cases grow quasi-exponentially while there are susceptible members of the population^[subject to the effectiveness of mitigation measures and testing availability], it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time. In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart. As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public. Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so with emotionally charged content is difficult.
% Discussion of how this section's conclusions change with the pandemic

@@

However, after the first wave of COVID, the issues with log scales became more apparent; \Cref{fig:log-scale-failures} demonstrates the problem. With a baseline number of cases that occur either because charts show a cumulative case count or because there is some baseline spread level that is not exponential because of mitigation measures ($R_{eff} \leq 1$), a log scale makes it difficult to detect slight increases in case counts that might indicate a new source of uncontrolled spread.

\Cref{fig:log-scale-failures} shows

% Problems with log scales

<<log-scale-failures>>=
# ----- Set up labeller to tweak things ----------------------------------------
rlistmap <- function(x, myfun) {
  if ("list" %in% class(x)) return(purrr::map(x, ~rlistmap(., myfun)))

@@

# ----- Set up data ------------------------------------------------------------
growth <- rpois(30, 15*exp((1:30 - 20)/5))

@@

# ----- Set up scales ----------------------------------------------------------
library(facetscales)
scales_y <- list(
  Log = scale_y_log10(),
  Linear = scale_y_continuous()
)

tmp %>%
ggplot(aes(x = x, y = value, color = name, size = name)) + 
  geom_line(alpha = .9) +
  facet_grid_sc(rows = vars(scale), cols = vars(Baseline), 
                scales = list(x = "fixed", y = scales_y)) +
  scale_size_manual(values = c(1.75, 1, 1)) +
  scale_color_manual(values = c("#249ab5", "#d00000", "#001226")) +
  scale_fill_discrete(name = "") + 
  theme(legend.position = "bottom", axis.title = element_blank(),
        legend.title = element_blank())
@

% Problems with linear scales

@@

@@

\bibliography{references}

>>>>

\usepackage{cleveref}

@@

\begin{document}
% \SweaveOpts{concordance=TRUE}
<<include=FALSE, cache=FALSE, echo=FALSE, purl=FALSE>>=
library(knitr)
opts_chunk$set(comment = NA, size = 'footnotesize', prompt = TRUE, highlight = FALSE, 
               cache = TRUE, crop = TRUE, concordance = FALSE, fig.align='center',
               fig.path='Figures_R/fig-',out.width="0.4\\textwidth", 
               message = F, error = F, warning = F, echo = F)
opts_knit$set(out.format = "latex")
knit_hooks$set(crop = hook_pdfcrop)
options(width = 80, prompt = "R> ", continue = "    ", digits = 3)
thm <- knit_theme$get('default')
Sys.setenv(TEXINPUTS=getwd(),
           BIBINPUTS=getwd(),
           BSTINPUTS=getwd())
knit_theme$set(thm)
@

@@

<<freq-domain-plot, include = F, eval = F>>=
show_states <- c("California", "Texas", "New York", "Arizona", "Iowa", "Washington")

@@

\section{To log or not to log}
As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time. In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart. As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public. Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult. When panic, fear, uncertainty, and doubt about the situation are added to the mix, 
% Discussion of how this section's conclusions change with the pandemic

@@

<<log-scale-initial, fig.width = 8, fig.height = 4, out.width = "\\linewidth", fig.cap = "In the early stages of the pandemic, log scales allowed the comparison of raw case counts in locations with vastly different population and case counts.">>=
# remotes::install_github("AllanCameron/geomtextpath")
library(geomtextpath)

library(facetscales)
scales_y <- list(
  Log = scale_y_log10(),
  Linear = scale_y_continuous()
)
tmp <- cvstate %>% 
  filter(date < ymd("2020-06-01")) %>%
  filter(state %in% c("Michigan", "Washington", "New York", "California"))

p1 <-  tmp %>%
  ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
  geom_textline(aes(color = state, label = state, vjust = -0.1, hjust =.93) , text_smoothing = 50) + 
  ylab("Daily average cases (7 day average)") + 
  xlab("Date") + 
  guides(color = "none") + 
  scale_y_log10() + 
  facet_grid(~"Log") + 
  theme_bw()

p2 <-  tmp %>%
  ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
  geom_textline(aes(color = state, label = state, vjust = -0.25, hjust = .85) , text_smoothing = 60) + 
  ylab("Daily average cases (7 day average)") + 
  xlab("Date") + 
  guides(color = "none") + 
  facet_grid(~"Linear") + 
  theme_bw() +
  theme(axis.title.y = element_blank())

grid.arrange(p1, p2, nrow = 1)
@

While log scales are not necessarily intuitive, many outlets tried to make the graphs more intuitive by adding reference lines, as shown in \Cref{fig:diag-ref-lines}.

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{ft-covid-19-deaths}
\caption{Reference lines to compare exponential growth rates of deaths in different countries. This provides some additional context that may help individuals use log scale data more successfully. This approach was first featured in the Financial Times, but was quickly adopted by the New York Times, 91-DIVOC, and other outlets. Graph from the Financial Times (March 23, 2020), image from \citet{kosaraPraiseDiagonalReference2020}.}
\label{fig:diag-ref-lines}
\end{figure}

% Problems with log scales
However, after the first wave of COVID, the issues with log scales became more apparent: it was difficult to detect slight increases in case counts that indicated the beginning of a new wave amid a background level of spread, as demonstrated in  \Cref{fig:log-scale-failures}. Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave.

<<log-scale-failures, echo = F, fig.width = 8, fig.height = 4, out.width = ".95\\linewidth", message = F, error = F, warning = F, fig.cap = "One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread. Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background. In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\\% of the cases at the end of the time period shown." >>=
# Set up labeller to tweak things 
rlistmap <- function(x, myfun) {
  if ("list" %in% class(x)) return(purrr::map(x, ~rlistmap(., myfun)))

@@

# Set up data 
growth <- rpois(30, 15*exp((1:30 - 20)/5))

@@

# Set up scales 
scales_y <- list(
  Log = scale_y_log10(),
  Linear = scale_y_continuous()
)

tmp %>%
ggplot(aes(x = x, y = value, color = name, size = name)) + 
  geom_line(alpha = .9) +
  facet_grid_sc(rows = vars(scale), cols = vars(Baseline), 
                scales = list(x = "fixed", y = scales_y)) +
  scale_size_manual(values = c(1.75, 1, 1)) +
  scale_color_manual(values = c("#249ab5", "#d00000", "#001226")) +
  scale_fill_discrete(name = "") + 
  theme_bw() + 
  theme(legend.position = "bottom", axis.title = element_blank(),
        legend.title = element_blank())
@

% Problems with linear scales
While log scales have their problems, linear scales are not immune from issues either. It can be very difficult to adequately compare to past situations when looking at the full time series of case counts. For example, in \Cref{fig:linear-scales-ref-lines}, it is difficult to tell whether the first wave of COVID cases in March 2020 had an increase as fast as that in January of 2021; it is even more difficult to compare the order-of-magnitude of change in case rate growth of January 2021 relative to January 2022 when the more contagious omicron variant became prevalent.

<<linear-scales-ref-lines, fig.width = 8, fig.height = 4, out.width = ".8\\linewidth", fig.cap = "Reported COVID cases in New York State, 2020-2022. The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \\emph{rate}. A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information.">>=
cvstate %>%
  filter(state == "New York") %>%
  ggplot(aes(x = date, y = daily_avg_cases)) + 
  geom_line() + 
  ggtitle("Reported COVID cases in New York State, 2020-2022") +
  ylab("Daily Cases reported (7-day average)") + 
  xlab("Date") + theme_bw()
@

<<log-scales-ref-lines, fig.width = 8, fig.height = 4, out.width = ".8\\linewidth", fig.cap = "Reported COVID cases in New York State, 2020-2022. The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past.">>=
cvstate %>%
  filter(state == "New York") %>%
  ggplot(aes(x = date, y = daily_avg_cases)) + 
  geom_line() + 
  ggtitle("Reported COVID cases in New York State, 2020-2022") +
  ylab("Daily Cases reported (7-day average)") + 
  xlab("Date") + theme_bw() + 
  scale_y_log10()
@

@@

It is not clear that the use of log or linear scales during the COVID-19 pandemic had a large effect on public opinion. Several studies were conducted in the early stages of the pandemic \citep{romanoScaleCOVID19Graphs2020, seviLogarithmicLinearVisualizations2020, ryanGraphsLogarithmicAxes2020} and results seem to suggest that while individuals have difficulty understanding log scale graphs, these issues do not tend to affect their support for intervention measures (perhaps in part because COVID-related news saturated the news and opinions were set outside of information provided in the experiments).

@@

\bibliography{refs}
